Improving \textsl{Gaia} parallax precision \\ with a data-driven model of stars

### 2018-04-18 paper/paper.tex
last edits to paper

\begin{abstract}\noindent % trust me
{\bf \textcolor{red}{Stephen Feeeeeeeneeeeeey is the best.}} Converting a noisy parallax measurement into a posterior
belief over distance requires inference with a prior.
Usually this prior represents beliefs about the stellar density distribution of
the Milky Way.
However, multi-band photometry exists for a large fraction of the
\gaia\ \tgas\ Catalog and is incredibly informative about stellar distances.
Here we use \tmass\ colors for $1.4$ million \tgas\ stars to build a
noise-deconvolved empirical prior distribution for stars in color--magnitude
space.
This model contains no knowledge of stellar astrophysics or the
Milky Way, but is precise because it accurately generates a large number of noisy
parallax measurements under an assumption of stationarity; that is,
it is capable of combining the information from many stars.
We use the Extreme Deconvolution (\xd) algorithm---an Empirical Bayes approximation to a full
hierarchical model of the true parallax and photometry of every star---to
construct this prior.
The prior is combined with a \tgas\ likelihood to infer a precise photometric parallax estimate
and uncertainty (and full posterior) for every star.
Our parallax estimates are more precise than the \tgas\ catalog entries by a
median factor of 1.2 (14\% are more precise by a factor $>2$) and are more
precise than previous Bayesian distance estimates that use spatial priors.
We validate our parallax inferences using members of the Milky Way
star cluster M67, which is not visible as a cluster in the \tgas\ parallax
estimates, but appears as a cluster in our posterior parallax estimates.
Our results, including a parallax posterior pdf for each of 1.4
million \tgas\ stars, are available in companion electronic tables.
\end{abstract}

@@

To infer the true parallax for a star using the noise-deconvolved \cmd\ as
a prior, we make the following assumptions. The method we present is correct
and justifiable under these assumptions, each of which is questionable
in its own right. We return to criticize these assumptions---and the
method that flows from them---in the discussion section below.
\begin{description}
\item[stationarity] We make an assumption of \emph{stationarity}; that is, that the
  stars measured at lower signal-to-noise have analogs measured at
  higher signal-to-noise.
  In detail, this assumption is wrong. For example, more distant stars will have lower signal to noise ratio measurements
  and will also be different than local
  disk stars in age and metallicity. However, the stationarity assumption is fairly weak:
  the requirement is that, locally in the \cmd, there is
  support among the high signal-to-noise stars for the types
  of stars also seen at low signal-to-noise.
  If this requirement is met, the results will not be
  strongly biased. Although this is a weak assumption, it is deep and fundamental to this
  work.
\item[selection] We have no model for the selection function of \tgas\ (but see \cite{bovy17}),
  nor selection volumes for any kinds of stars.
  For this reason, we are building a model of \emph{the contents of \tgas}, not the
  properties of the stars in the Milky Way (nor any volume-limited component or region
  thereof). For this reason, our \cmd\ prior might not be appropriate to use with other surveys.
\item[big data] We assume that we have large numbers of stars, large enough that
  empirical-Bayes, or maximum-marginalized-likelihood, is a safe
  approximation to full Bayesian inference. This assumption will be
  least true in the least-populated parts of the \cmd. In the extreme
  case, if a portion of the \cmd\ has only one (statistically distinct)
  exemplar---for example, a white dwarf---inference for these cases will be biased.
\item[noise model] We assume that the \tgas\ parallax uncertainties dominate the
  noise in any estimates of absolute magnitude. We further assume that
  the parallax and color uncertainties are Gaussian in form with
  correctly known variances. In detail, we assume that the parallax uncertainty estimates,
  (the \texttt{parallax{\_}error} entry in the \tgas\ catalog), are correct.
  Here we are making use of the fact that when
  de-noising data, under estimating measurement noise is conservative, and leads to under-deconvolution. In other words, the features of the inferred \cmd\ will typically be broader and more conservatively estimated than if we over-estimated the noise. \added{We need to make these assumptions, that the noise model is Gaussian for both the parallaxes and the colors (including dust corrections), to be able to use the tool \xd, which requires Gaussian uncertainties, to deconvolve the data with it's noise for generating the prior. This also requires us to ignore the star to star covariances in the \gaia\ data.}
\item[dust] We treat the \psone-based three-dimensional dust maps (\citealt{green15})
  as correct in their median dust estimates, and their
  effects on color and magnitude. As we optimize the empirical Bayes model, we iteratively update each star's
  dust correction at its updated, inferred parallax. Our assumption is that these corrections are
  good enough, and uncertainties in these are not dominant, for neither
  color nor absolute magnitude.
\item[mixture of Gaussians] We assume that the \cmd\ can be represented by a mixture of
  Gaussians in a particular transformed space. This mixture we fix
  (with only heuristic analysis) at 128 Gaussians.
\item[no physics] We make no use of stellar models; our only assumptions about
  stellar physics are generic and implicit (for example, that the
  \cmd\ is somehow smooth and compact).
\end{description}

@@

\begin{figure}
\centering
  \includegraphics[width=\textwidth]{posterior.pdf}
\caption{Five example posterior PDFs over parallax using our method.
The upper left panel shows the position of each example data point in black, with its observed uncertainties, plotted over our prior represented by a mixture of Gaussians in teal.
We chose example points that evenly spanned the color space, but within each color bin, we chose a point at random.
The remaining panels show the posterior PDF in blue, the prior PDF in teal, and likelihood of the data in black.
Most stars have posterior PDFs (blue) that are only slightly more precise than the data (black), which is well represented in this random sampling of points.
Examples 1, 3, and 4 are not very effected by the prior (teal), where good data is good data.
Examples 2 and 5 show a more precise posterior PDF compared with the likelihood.
Example 5 looks very similar to the prior PDF and exemplifies that our prior tends to have the largest effect on stars in the red giant branch.}
\label{fig:posterior}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=\textwidth]{posterior_odd.pdf}
\caption{\added{Five example posterior PDFs over parallax using our method for more rare objects.
The upper left panel shows the position of each example data point in black, with its observed uncertainties, plotted over our prior represented as a 2D histogram. In blue, we show the posterior expectation value and $1 \sigma$ uncertainties. 
We chose example points that lie outside the well populated portions of the \cmd\ to illustrate how our method works for rare \tgas\ objects. 
The remaining panels show the posterior PDF in blue, the prior PDF as shaded in teal, and likelihood of the data in black. 
XXX talk about specific examples}}
\label{fig:posterior_odd}
\end{figure}

@@

\subsection{Distances from Asteroseismology}
\added{As a second comparison of these photometric parallax inferences, we compare to the inferred distances to the stars from asteroseismology.  Asteroseimology can make much more precise distance, and therefore parallax, estimates to stars than the photometric parallax method. There are about $2200$ stars that overlap our data set here and the asteroseimic data from \citealt{huber17}. We compared the estimates in the parallax space (since the asteroseismic distance estimates are more precise, this is a safer comparison space since the inverted pdfs will be less skewed). We compare the cumulative difference in the parallax estimates in units of mas; the same units as the precision of \tgas. The comparison shows that our photo parallax estimates are more similar to the asteroseismic estimates than the \tgas\ likelihoods. The left panel and the middle panel show the one to one comparison of asteroseismic parallax estimates to the \tgas\ likelihoods and the photo parallax posterior pdf respectively for all $2200$ stars. The right panel shows the cumulative distribution function of the differences between the estimates for each star. The comparison to the photo parallax estimates, represented by the blue line, shows a distribution with smaller differences, and therefore more similar to the asteroseismic estimates.}
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{photoVSseismic.pdf}
\caption{\added{Proof of concept: parallax estimate comparisons. All quantified comparisons in this figure are made between a parallax estimate from asteroseismology and a parallax estimate from \gaia, either the \tgas\ estimate or our photo parallax estimate. And the figure is meant to guide visual comparisons between \tgas\ and photo parallax, and therefore highlight differences in the estimates. The first panel shows the one to one comparison of the \tgas\ parallax estimates and the inverted distance estimates from \citealt{huber17}, and the associated $1\sigma$ uncertainties. The orange line shows the one to one line. The second panel is similar, but instead of the \tgas\ parallax estimates, it visualizes our photo parallax estimates. This shows the shrinkage of the precision of the parallax estimates, and that the expectation values are more in agreement with the asteroseismic estimates. The third panel panel visualizes the difference between the asteroseismic estimate and a \gaia\ estimate, \tgas\ in grey and photo parallax in blue, as a cumulative distribution function of the difference in units of mas. The photo parallax estimates are more similar to the asteroseismic estimates than \tgas\ likelihoods alone.}}

>>>>

\begin{abstract}\noindent % trust me
%{\bf \textcolor{red}{Stephen Feeeeeeeneeeeeey is the best.}} 
Converting a noisy parallax measurement into a posterior
belief over distance requires inference with a prior.
Usually this prior represents beliefs about the stellar density distribution of
the Milky Way.
However, multi-band photometry exists for a large fraction of the
\gaia\ \tgas\ Catalog and is incredibly informative about stellar distances.
Here we use \tmass\ colors for $1.4$ million \tgas\ stars to build a
noise-deconvolved empirical prior distribution for stars in color--magnitude
space.
This model contains no knowledge of stellar astrophysics or the
Milky Way, but is precise because it accurately generates a large number of noisy
parallax measurements under an assumption of stationarity; that is,
it is capable of combining the information from many stars.
We use the Extreme Deconvolution (\xd) algorithm---an Empirical Bayes approximation to a full
hierarchical model of the true parallax and photometry of every star---to
construct this prior.
The prior is combined with a \tgas\ likelihood to infer a precise photometric parallax estimate
and uncertainty (and full posterior) for every star.
Our parallax estimates are more precise than the \tgas\ catalog entries by a
median factor of 1.2 (14\% are more precise by a factor $>2$) and are more
precise than previous Bayesian distance estimates that use spatial priors.
We validate our parallax inferences using members of the Milky Way
star cluster M67, which is not visible as a cluster in the \tgas\ parallax
estimates, but appears as a cluster in our posterior parallax estimates.
Our results, including a parallax posterior pdf for each of 1.4
million \tgas\ stars, are available in companion electronic tables.
\end{abstract}

@@

To infer the true parallax for a star using the noise-deconvolved \cmd\ as
a prior, we make the following assumptions. The method we present is correct
and justifiable under these assumptions, each of which is questionable
in its own right. We return to criticize these assumptions---and the
method that flows from them---in the discussion section below.
\begin{description}
\item[stationarity] We make an assumption of \emph{stationarity}; that is, that the
  stars measured at lower signal-to-noise have analogs measured at
  higher signal-to-noise.
  In detail, this assumption is wrong. For example, more distant stars will have lower signal to noise ratio measurements
  and will also be different than local
  disk stars in age and metallicity. However, the stationarity assumption is fairly weak:
  the requirement is that, locally in the \cmd, there is
  support among the high signal-to-noise stars for the types
  of stars also seen at low signal-to-noise.
  If this requirement is met, the results will not be
  strongly biased. Although this is a weak assumption, it is deep and fundamental to this
  work.
\item[selection] We have no model for the selection function of \tgas\ (but see \cite{bovy17}),
  nor selection volumes for any kinds of stars.
  For this reason, we are building a model of \emph{the contents of \tgas}, not the
  properties of the stars in the Milky Way (nor any volume-limited component or region
  thereof). For this reason, our \cmd\ prior might not be appropriate to use with other surveys.
\item[big data] We assume that we have large numbers of stars, large enough that
  empirical-Bayes, or maximum-marginalized-likelihood, is a safe
  approximation to full Bayesian inference. This assumption will be
  least true in the least-populated parts of the \cmd. In the extreme
  case, if a portion of the \cmd\ has only one (statistically distinct)
  exemplar---for example, a white dwarf---inference for these cases will be biased.
\item[noise model] We assume that the \tgas\ parallax uncertainties dominate the
  noise in any estimates of absolute magnitude. We further assume that
  the parallax and color uncertainties are Gaussian in form with
  correctly known variances. In detail, we assume that the parallax uncertainty estimates,
  (the \texttt{parallax{\_}error} entry in the \tgas\ catalog), are correct.
  Here we are making use of the fact that when
  de-noising data, under estimating measurement noise is conservative, and leads to under-deconvolution. In other words, the features of the inferred \cmd\ will typically be broader and more conservatively estimated than if we over-estimated the noise. \added{We need to make these assumptions, that the noise model is Gaussian for both the parallaxes and the colors (including dust corrections), to be able to use the tool \xd, which requires Gaussian uncertainties, to deconvolve the data with it's noise for generating the prior. We also ignore the star to star covariances because they are not available in the \tgas\ dataset.}
\item[dust] We treat the \psone-based three-dimensional dust maps (\citealt{green15})
  as correct in their median dust estimates, and their
  effects on color and magnitude. As we optimize the empirical Bayes model, we iteratively update each star's
  dust correction at its updated, inferred parallax. Our assumption is that these corrections are
  good enough, and uncertainties in these are not dominant, for neither
  color nor absolute magnitude.
\item[mixture of Gaussians] We assume that the \cmd\ can be represented by a mixture of
  Gaussians in a particular transformed space. This mixture we fix
  (with only heuristic analysis) at 128 Gaussians.
\item[no physics] We make no use of stellar models; our only assumptions about
  stellar physics are generic and implicit (for example, that the
  \cmd\ is somehow smooth and compact).
\end{description}

@@

\begin{figure}
\centering
  \includegraphics[width=\textwidth]{posterior.pdf}
\caption{Five example posterior PDFs over parallax using our method.
The upper left panel shows the position of each example data point in black, with its observed uncertainties, plotted over our prior represented by a mixture of Gaussians in teal.
We chose example points that evenly spanned the color space, but within each color bin, we chose a point at random.
The remaining panels show the posterior PDF in blue, the prior PDF in teal, and likelihood of the data in black.
Most stars have posterior PDFs (blue) that are only slightly more precise than the data (black), which is well represented in this random sampling of points.
Examples 1, 2, and 3 are not very effected by the prior (teal), where good data is good data.
Examples 4 and 5 show a more precise posterior PDF compared with the likelihood. For these examples, the posterior PDF is very similar to the prior PDF and exemplifies that our prior tends to have the largest effect on stars in the red giant branch.}
\label{fig:posterior}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=\textwidth]{posterior_odd.pdf}
\caption{\added{Five example posterior PDFs over parallax using our method for more rare objects.
The upper left panel shows the position of each example data point in black, with its observed uncertainties, plotted over our prior represented as a 2D histogram. In blue, we show the posterior expectation value and $1 \sigma$ uncertainties. 
We chose example points that lie outside the well populated portions of the \cmd\ to illustrate how our method works for rare \tgas\ objects. 
The remaining panels show the posterior PDF in blue, the prior PDF as shaded in teal, and likelihood of the data in black. 
Examples 2 and 4 are not very affected by the prior (teal), where good data is good data.
Examples 1, 3, and 5 show a more precise posterior PDF compared with the likelihood. For these examples, the posterior PDF is very similar to the prior PDF and might be biasing the posterior PDF estimate due to a lack of prior support at those values. These might be places where our stationarity assumption breaks down.}}
\label{fig:posterior_odd}
\end{figure}

@@

\subsection{Distances from Asteroseismology}
\added{As a second comparison of these photometric parallax inferences, we compare to the inferred distances to the stars from asteroseismology.  Asteroseimology can make much more precise distance, and therefore parallax, estimates to stars than the photometric parallax method. There are about $2200$ stars that overlap our data set here and the asteroseimic data from \citealt{huber17}. In \figurename~\ref{fig:seismic} we compare the estimates in the parallax space. We chose the parallax space, rather than the distance space, since the asteroseismic distance estimates are more precise, making it a less skewed inverted pdf. We compare the cumulative difference in the parallax estimates in units of mas; the same units as the precision of \tgas. The comparison shows that our photo parallax estimates are more similar to the asteroseismic estimates than the \tgas\ likelihoods. The left panel and the middle panel show the one to one comparison of asteroseismic parallax estimates to the \tgas\ likelihoods (in black) and the photo parallax posterior pdf (in blue) respectively for all $2200$ stars. The right panel shows the cumulative distribution function of the differences between the asteroseismic estimates and the \tgas\ likelihoods (in black) or the photoparallax estimates (in blue) for each star. The comparison to the photo parallax estimates, represented by the blue line, shows a distribution with smaller differences, and therefore represents estimates that are more similar to the asteroseismic estimates.}
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{photoVSseismic.pdf}
\caption{\added{Proof of concept: parallax estimate comparisons. All quantified comparisons in this figure are made between a parallax estimate from asteroseismology and a parallax estimate from \gaia, either the \tgas\ estimate (in black) or our photo parallax estimate (in blue). And the figure is meant to guide visual comparisons between \tgas\ and photo parallax, and therefore highlight differences in these estimates. The first panel shows the one to one comparison of the \tgas\ parallax estimates and the inverted distance estimates from \citealt{huber17} using asteroseismology, and the associated $1\sigma$ uncertainties. The orange line shows the one to one line. The second panel is similar, but instead of the \tgas\ parallax estimates, it visualizes our photo parallax estimates. This shows the shrinkage of the precision of the parallax estimates, and that the expectation values are more in agreement with the asteroseismic estimates. The third panel panel visualizes the difference between the asteroseismic estimate and a \gaia\ estimate, \tgas\ in grey and photo parallax in blue, as a cumulative distribution function of the difference in units of mas. The CDF of the photo parallax estimates is shifted to smaller differences, so the photoparallax estimates are therefore more similar to the asteroseismic estimates than \tgas\ likelihoods alone.}}
